HO unification from object language to meta language

### 2024-02-16 main.tex
add todo in tex
update tex

  % Unification is a fundamental process behind goal resolution in logic
  % programming. Two terms $t_1$ and $t_2$ unify up to $\eta\beta$-reduction if it
  % exist a substitution $\theta$ such that $\theta t_1 = \theta t_2$. A
  % substitution is a mapping from unification variables to terms and it is the most
  % general, if forall substitution $\theta'$, such that $\theta' t_1 = \theta'
  % t_2$, it exists a substitution $\theta''$ such that $\theta' = \theta \theta''$.
  % Unification has been widely studied in the 90s [here the refs]. In this paper we
  % want to give a framework to reason about unification in meta-programming.
  Meta-programming \cite{hill1994} is a programming technique in which a
  program, referred to as meta language (\ml), can treat an other program, knwon
  as the object language (\ol) as its data. This programming approach is used in
  various works \cite{cite1, cite2...} to construct logical frameworks (\lf)
  capable of automating proof resolution. In many of the previously the
  previously mentioned works, the equational theory expressed by the \ol is
  weaker than the \ml one. Consequently, no effort is required to adjust the
  unification algorithm of the \ml since it encompass the \ml one. In our
  specific scenario, however, we work with an \ol with a richer equational
  theory than our \ml. This paper introduces a framework that enable the \ml to
  comprehend the unification of the \ol and provides some practical applications
  of these findings. Additionally, we extend our work to unification in the case
  of higher-order variable quantification.

@@

% theorem prover and tc
Thanks to meta programming, it is feasible to express the equational theory of a
theorem prover, as indicated in \cite{citeX, Y Z}, with the \ml serving as the
foundational infrastructure for this ecosystem. The process of proving a theorem, that is
solving a query using the jargon of logic programming consist in verifying if a
proposition is a logical consequence of a given set of assumption. The benefit
of of using a prologish verification style is that the proof resolution can be
automatized thanks to different search strategies, such as a depth first search
with backtracking. One prominent manifestation of this automated proof process is
type-class resolution\cite{wadler89, sozeau08}. Type classes serve as a typing
structure to introduce ad-hoc polymorphism in functional languages.
In \coq, the use of type class has increasingly become a programming
style. Numerous libraries, such as \stdpp and
\iris are built upon this machinery.

@@

\noindent where a \coq constant is a gref\footnote{a gref is the opaque type for \coq
identifier in \elpi} inside the \elpiIn{const} constructor, the \coq application
is embedded in the \elpiIn{app} node, having the head and the arguments of the
application expressed as a list of terms. Lambda abstractions are translated
into the node \elpiIn{fun} binding a term to another term. Finally forall
quantification on the for \coqIn{~$\forall$~x,F} into \elpiIn{prod x\ F}.

@@

\noindent In our type-class solver, we translate the \coqIn{forall_dec} instance
of \coq (on the left) into the \elpi rule (on the right). This translation, derived
directly from the type of \coqIn{forall_dec}, conceals certain unification
properties that are accepted by the unification algorithm of the \ol but 
rejected by the unification algorithm of the \ml. The primary issue of this
translation lies in how the subterm \coqIn{~$\forall$~x, P x} is rendered in
the \ml rule; \coqIn{P} is a higher-order variable, and in the original subterm, it
sees the binder \coqIn{x}. However, the rule in the \ml interprets \elpiIn{P} as a 
first-order variable, not seeing \elpiIn{x}. Additionally, this subterm is
nested within the \elpiIn{app} constructor, introducing a structural challenge 
to the unification process.

@@

Based on the definition provided earlier, the terms \elpiIn{app[const `f, const
`y, x]} and \elpiIn{app[P, x]} are expected to unify with substitution $\theta := \{P
\mapsto \texttt{\elpiIn{app [const `f, const `y]}} \}$. However this unification
is not possible in the \ml. This difficulty arises because, in this instance,
we have two terms with the same head
(the \elpiIn{app} node), but their corresponding lists have different lengths.

@@

\noindent In this second implementation, any call to the \elpiIn{tc-Decision}
will unify with the head of the rule for \coqIn{forall_dec}. The first
premise ensures that the arguments unify with what we expect from the
instance definition. Notably, in this case, the unification algorithm of
the \ol is leveraged, , successfully addressing the previously mentioned unification problems.

However, this approach has two main drawbacks. Firstly, the communication
between the \ol and the \ml may potentially be slow due to the transformation of
terms from one language to the other \todo{here I refer to constr->elpi and
elpi->constr}. Secondly, the alternative encoding lacks the ability to
discriminate clauses based on the head of the clause. For example, any indexing
algorithm designed to filter rules becomes ineffective in this scenario.

@@

% da notare che il problema dopo, is the FO
% approx. Coq applica la regola forall\_dec sul goal (prod x\ app[f, y, x]), ma forse 
% bisognerebbe vedere se altri solver accettano questa euristica. Così facciamo 
% vedere che non siamo i soli a farlo. Nel caso contrario bisogna dire come funziona
% l'algo di unif di coq in questo caso specifico

% where \coqIn{f : A -> B -> Type} and \coqIn{y: A}, we can notice that the \elpi
% rule above cannot be applied; the terms \elpiIn{prod x\ app[P, x]} and
% \elpiIn{prod x\ app[f, y, x]} cannot unify since the lists under the
% \elpiIn{app} constructors are of different length.\todo{NOTE: the goal unify
% with forall\_dec in coq xké fa unif FO approx}

% Another example of unification failure is given by the the following instance

>>>>

  % Unification is a fundamental process behind goal resolution in logic
  % programming. Two terms $t_1$ and $t_2$ unify up to $\eta\beta$-reduction if it
  % exist a substitution $\theta$ such that $\theta t_1 = \theta t_2$. A
  % substitution is a mapping from unification variables to terms and it is the most
  % general, if forall substitution $\theta'$, such that $\theta' t_1 = \theta'
  % t_2$, it exists a substitution $\theta''$ such that $\theta' = \theta \theta''$.
  % Unification has been widely studied in the 90s [here the refs]. In this paper we
  % want to give a framework to reason about unification in meta-programming.
  Meta-programming \cite{hill1994} is a programming technique in which a
  program, referred to as meta language (\ml), can treat an other program, knwon
  as the object language (\ol) as its data. This programming approach is used in
  various works \cite{cite1, cite2...} to construct logical frameworks (\lf)
  capable of automating proof resolution. In many of the
  previously mentioned works, the equational theory expressed by the \ol is
  weaker than the \ml one. Consequently, no effort is required to adjust the
  unification algorithm of the \ml to represent the unification process of 
  the \ol: the former algorithm encompassing the second one. In our
  specific scenario, however, we work with an \ol with a richer equational
  theory than our \ml. This paper introduces a framework that enables the \ml to
  comprehend the unification of the \ol in the context of higher-order 
  unification. Practical applications of these findings will be provided.

@@

% theorem prover and tc
Thanks to meta programming, it is feasible to express the equational theory of a
theorem prover, as indicated in \cite{X, Y, Z}, with the \ml serving as the
foundational infrastructure for this ecosystem. The process of proving a theorem, that is
solving a query using the jargon of logic programming, consist in verifying if a
proposition is a logical consequence of a given set of assumption. The benefit
of using a prologish verification style is that the proof resolution can be
automatized thanks to different search strategies, such as a depth first search
with backtracking. One prominent manifestation of this automated proof process is
type-class resolution\cite{wadler89, sozeau08}. Type classes serve as a typing
structure to introduce ad-hoc polymorphism in functional languages.
In \coq, the use of type classes has increasingly become a programming
style. Numerous libraries, such as \stdpp and
\iris are built upon this machinery.

@@

\noindent where a \coq constant is a gref\footnote{a gref is the opaque type for \coq
identifier in \elpi} inside the \elpiIn{const} constructor, the \coq application
is embedded in the \elpiIn{app} node, having the head and the arguments of the
application expressed as a list of terms. Lambda abstractions are translated
into the node \elpiIn{fun}, binding a term to another term. Finally forall
quantification on the form \coqIn{~$\forall$~x,F} in the \ol are translated
into \elpiIn{prod x\ F} within the \ml.

@@

\noindent In our type-class solver, the \coqIn{forall_dec} instance 
of \coq (on the left) is compiled into the \elpi rule (on the right). This 
compilation, derived
directly from the type of \coqIn{forall_dec}, conceals certain unification
properties that are accepted by the unification algorithm of the \ol but 
rejected by the unification algorithm of the \ml. The primary issue of this
compilation lies in how the subterm \coqIn{~$\forall$~x, P x} is rendered in
the \ml rule: \coqIn{P} is a higher-order variable and it
sees the binder \coqIn{x}. However, the rule in the \ml interprets \elpiIn{P} as a 
first-order variable, not seeing \elpiIn{x}. Additionally, this subterm is
nested within the \elpiIn{app} constructor, introducing a structural challenge 
to the unification process.

@@

Based on the definition provided earlier, the terms \elpiIn{app[const `f, const
`y, x]} and \elpiIn{app[P, x]} are expected to unify with substitution $\theta := \{P
\mapsto \texttt{\elpiIn{app [const `f, const `y]}} \}$. However this unification
is not possible in the \ml. This difficulty arises because, even though 
the two terms share the same same head
(the \elpiIn{app} node), their corresponding lists have different lengths.

@@

\noindent In this second implementation, any query to \elpiIn{tc-Decision}
will unify with the head of the rule above. The first
premise ensures that the arguments unify with what we expect from the
instance definition. Notably, in this case, the unification algorithm of
the \ol is leveraged successfully addressing the previously mentioned unification problems.

However, this approach has two main drawbacks. Firstly, the communication
between the \ol and the \ml may potentially be slow due to the transformation of
terms from one language to the other \todo{here I refer to constr->elpi and
elpi->constr}. Secondly, a lack of % the ability to
clauses discrimination based on the head of the clause is no longer
possible. For example, any indexing
algorithm designed to filter rules becomes ineffective in this scenario.

@@

% TODO: parlare non so ancora dove dell'encoding shallow
% dei terming ol in ml. Invece di usare i nodi app, prod etc... riusiamo
% la stessa sintassi del linguaggio oggetto. Questo potrebbe funzionare 
% bene nel caso di un ol con logica equazionale piu debole del ml, ma nel nostro
% caso non si può applicare. inoltre vogliamo tenere un encoding più strutturato
% per meglio analizzare i termini che ci arrivano nel ml.

% TODO: related work: chiedere a enrico

% nuova section dove si da la spec dei due linguaggi

% TODO: spiegare le proprietà di unification
